Racial and Ethnic Differences in ETI

### 2024-02-28 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
kam eti update

---
title: "Racial and Ethnic Differences in ETI"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    number-sections: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: C:/Users/timvigers/Dropbox/Miscellaneous/zotero.bib
csl: C:/Users/timvigers/Dropbox/Miscellaneous/american-medical-association.csl
editor: source
---

@@

## GLI-2012 vs. GLI-2022

GLI-2012 spirometry equations include race/ethnicity in the formula, but GLI-2022 equations are race-free. The results of the GLI-2022 equations contain the "gl" suffix, which stands for "global" (so instead of "ppFEV1" it would be "ppFEV1gl").

First, I wanted to ensure that the `rspiro` package is calculating FEV1 percent predicted correctly. In the graph below, ppFEV1 from the registry is shown on the x-axis and ppFEV1 as calculated with `rspiro` is shown on the y-axis.

```{r}
#| warning: false
ggplot(encounter, aes(
  x = GLI_FEV1_pct_predicted, y = GLI_FEV1_pct_predicted_check,
  color = race
)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  theme_bw() +
  xlab("Registry") +
  ylab("rspiro") +
  ylim(0, 150) +
  facet_wrap(~race) +
  ggtitle("Registry values vs. rspiro")
```

```{r}
ggplot(encounter, aes(
  x = ppFEV1gl, y = GLI_FEV1_pct_predicted_check,
  color = race
)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  theme_bw() +
  xlab("GLI-2022") +
  ylab("GLI-2012") +
  ylim(0, 150) +
  xlim(0, 150) +
  facet_wrap(~race) +
  ggtitle("GLI-2022 vs. GLI-2012 (rspiro)")
```

@@

The most basic model we can fit to these data uses the average values as the outcome, with terms for review year, race, and their interaction and a random intercept for each patient.

```{r}
# Most basic model of lung function over time by race
# Try a few different models
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
label(means) <- list("Race/Ethnicity", "Year", "Mean ppFEV", "SE", "df", "Lower CL", "Upper CL")
gt(means) %>% fmt_number(emmean:last_col(), decimals = 1)
```

```{r}
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(
    x = ReviewYear, ymin = emmean - SE,
    ymax = emmean + SE, color = race
  ), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
#| warning: false
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(
    mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID),
    alpha = 0.005, linewidth = 0.5
  ) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(x = ReviewYear, y = emmean, group = race, color = race)) +
  geom_errorbar(mapping = aes(x = ReviewYear, ymin = emmean - SE, ymax = emmean + SE, color = race), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity")
```

```{r}
#| warning: false
ggplot(left_join(annual, means, by = join_by(race, ReviewYear))) +
  geom_line(
    mapping = aes(x = ReviewYear, y = mean_ppFEV, group = eDWID),
    alpha = 0.005, linewidth = 0.5
  ) +
  geom_point(mapping = aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line(mapping = aes(
    x = ReviewYear, y = emmean, group = race,
    color = race
  )) +
  geom_errorbar(mapping = aes(
    x = ReviewYear, ymin = emmean - SE,
    ymax = emmean + SE, color = race
  ), width = .2) +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  theme_bw() +
  scale_color_discrete(name = "Race/Ethnicity") +
  facet_wrap(~race)
```

#### Change point model

```{r}
# Calculate time from first ETI use
encounter <- encounter %>%
  mutate(
    days_from_start =
      as.numeric(encounterdate - Modulator_trikafta_first_date)
  )
# Segmented regression by group
lm_fixed_cp <- segreg(
  GLI_FEV1_pct_predicted ~ race + seg(days_from_start, by = race),
  data = encounter
)
# Results
lm_fixed_cp$psi
slope(lm_fixed_cp)
# Plot
new_df = lm_fixed_cp$model
new_df$pred = predict(lm_fixed_cp)
ggplot(new_df, aes(x = days_from_start, y = pred,group = race,color = race)) +
  geom_line() +
  ylim(0, 150)
# # Random intercept model isn't converging
# # Basic model
# lmm_base <- lme(GLI_FEV1_pct_predicted ~ days_from_start,
#   random = ~ 1 | eDWID, data = encounter, na.action = na.omit,method = "ML",
#   control=list(msMaxIter=1000,msMaxEval = 1000)
# )
# # Segmented with fixed change point
# lmm_fixed_cp <- segmented(lmm_base,
#   seg.Z = ~days_from_start, psi = 0,
#   random = list(eDWID = pdDiag(~ 1 + days_from_start + U))
# )
```

```{r}
#| cache: true
t <- encounter %>%
  select(eDWID, GLI_FEV1_pct_predicted, days_from_start) %>%
  drop_na()
model <- list(
  GLI_FEV1_pct_predicted ~ 0 + days_from_start,
  ~ 0 + days_from_start
)
fit <- mcp(model, data = t, cores = 4, chains = 4)
plot(fit)
```

>>>>

---
title: "Racial and Ethnic Differences in ETI"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    number-sections: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: /Users/timvigers/Documents/Miscellaneous/zotero.bib
csl: /Users/timvigers/Documents/Miscellaneous/american-medical-association.csl
editor: source
---

@@

@@

The most basic model we can fit to these data uses the average values as the outcome, with terms for review year, race, and their interaction and a random intercept for each patient. This model is adjusted for age, private vs. other insurance, and parental education.

```{r}
# Most basic model of lung function over time by race
# Try a few different models
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race | ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
```
